\mytitle

### 2021-10-18 paper/noms-short-paper.tex
Terrible, terrible things.

\begin{abstract}
Automatic, data-driven networking is becoming more capable and widely researched, partly driven by the efficacy of \emph{Deep Reinforcement Learning} (DRL) algorithms.
Yet the complexity of both DRL inference and learning force these tasks to be offloaded \emph{away from the dataplane} to hosts, harming latency-sensitive applications.
Online learning of such policies cannot occur in the dataplane, despite being useful techniques when problems evolve or are hard to model.

We present \emph{\approachshort{}}---\approach{}---the first work to bring \emph{online reinforcement learning} to the dataplane.
\approachshort{} makes online learning possible in constrained SmartNIC hardware by returning to classical RL techniques---avoiding neural networks.
Our design allows weak yet highly parallel SmartNIC NPUs to be competitive against commodity x86 hosts, despite having fewer features and slower cores.
Compared to offloading, we achieve a \SI{21}{\texttimes} reduction in \num{99.99}\nthscript{th} tail inference times to \SI{34}{\micro\second}, and \SI{9.9}{\texttimes} improvement in online throughput for real-world policy designs.
In-NIC execution eliminates PCIe transfers, and our asynchronous compute model ensures minimal impact on traffic carried by a co-hosted P4 dataplane.
\approachshort{}'s design scales with additional resources at compile-time to improve upon both decision latency and throughput, and are quickly reconfigurable at runtime compared to reinstalling device firmware.
\end{abstract}

@@

%\vspace{-1em}
\section{Introduction}
\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.65\linewidth]{figures/arch-with-p4}
	\caption{\approachshort{} brings low-latency, online reinforcement learning directly to the dataplane. SoC- and NetFPGA-based SmartNIC devices expose spare compute---making in-situ, asynchronous processing and learning possible alongside P4 dataplanes. Classical RL policy methods are the key to making this computationally feasible.\label{fig:netro-arch}}
\end{figure}
Automatic network optimisation, control, and defence are at last becoming commonplace.
Adaptive techniques such as \emph{reinforcement learning} (RL) have led the charge in data-driven networking, enhancing automatic traffic optimisation~\parencite{DBLP:conf/sigcomm/ChenL0L18,DBLP:conf/sigcomm/MaoNA17}, congestion control~\parencite{DBLP:journals/corr/abs-1910-04054}, adaptive routing~\parencite{DBLP:conf/hotnets/ValadarskySST17,DBLP:conf/conext/GiladSGRS20}, resource management~\parencite{DBLP:conf/hotnets/MaoAMK16}, and packet classification~\parencite{DBLP:conf/sigcomm/LiangZJS19}.
In RL methods, every change and its effects improve future decisions.

@@

In parallel, P4~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} and \emph{programmable dataplane} (PDP) hardware~\parencite{DBLP:journals/micro/ZilbermanACM14, netronome-smartnic, xilinx-alveo, barefoot-intel} have inspired explosive growth and interest in the research community surrounding in-network computation and offloading.
The promise of PDP hardware is that we can move the entire monitoring and analysis stack \emph{into the dataplane itself, and have it evolve to incorporate new approaches}.
%Consider a DDoS mitigation system which could react to dataplane-specific, per-packet state in microseconds, and \emph{start learning again just as quickly} in the face of new data or behaviours.
The P4 ecosystem already presents novel, openly-available, fine-grained traffic measurement techniques that can be installed and controlled with ease~\parencite{DBLP:conf/sigcomm/GuptaHCFRW18,DBLP:conf/sigcomm/ChenFKRR18,DBLP:conf/sosr/GhasemiBR17}.
In addition, P4's control plane makes it easy to select which flows or packets are monitored in a live network and potentially allow control over traffic at the decision site.
As a result, there has been keen interest in executing ML in the dataplane~\parencite{DBLP:conf/hotnets/XiongZ19,DBLP:conf/sigcomm/SanvitoSB18,DBLP:journals/corr/abs-1801-05731,DBLP:journals/corr/abs-2009-02353,langlet-ml-netronome,DBLP:journals/corr/abs-2002-08987} to take advantage of flow or per-packet state that cannot be efficiently processed or extracted at any other location in the network.
These works have shown the value of in-network ML: high-throughput, low latency response to network changes.
While they can exploit on-device state to provide reactive insight, the missing piece of the puzzle is learning and updating these ML analyses online without deferring to another machine in the network.
Training these models online and in-network is an exciting (and challenging) lacuna in the field that \emph{has yet to be addressed by the community}.

It is important to make this feasible; offloading to commodity hosts adds PCIe delays~\parencite{DBLP:journals/corr/abs-2009-02353,DBLP:conf/sigcomm/NeugebauerAZAL018}, but is required due to the complexity of modern ML.
For context, deep neural network training relies on backpropagation, can take vast amounts of offline simulation for RL~\parencite{DBLP:journals/corr/abs-1912-06680}, and needs many minibatches for stability.
These lead to high compute and storage costs, inducing capital expenditure for dedicated accelerators to overcome the batching needed to operate at line-rate---while BrainWave~\parencite{DBLP:conf/isca/FowersOPMLLAHAG18} can reduce batching (and thus tail latencies) by \qty{32}{\texttimes}~\parencite{Duarte2019}, inference still takes $\mathcal{O}{\left(\text{\si{\milli\second}}\right)}$.
Moreover, novel DMA techniques such as \emph{GPUDirect}~\parencite{gpudirect} halve but do not eliminate such PCIe transfers.
The high latencies \emph{caused} by offloading and inference harm learning~\parencite{DBLP:journals/firai/TravnikMSP18} and runtime application performance~\parencite{DBLP:journals/corr/abs-1910-04054}.
If we can bring online learning \emph{to the dataplane}, then we can take advantage of rich, local state while minimising these latencies (and their impact on the learned policy).
This would also make it easy to train and prototype agent designs which can learn \emph{as the environment evolves}, or when there is too little data to model and simulate a problem.

@@

In this paper, we enable \emph{online in-NIC learning} by returning to \emph{classical} RL methods.
In particular, we focus on tile coding with one-step temporal-difference algorithms such as Sarsa~\parencite{RL2E}.
%These choices have important benefits for in-NIC execution.
While these functions have lower model capacity, they do not require batches of inputs to learn in a stable way, negating the memory needed to store experience replays, and have simple update and inference logic.
In addition, they have shown promise in other network use cases~\parencite{DBLP:journals/eaai/MalialisK15,DBLP:journals/tnsm/SimpsonRP20}.
Tile coding in particular admits many optimisations, being an embarrassingly parallel problem.
Using fixed-point arithmetic, we solve the lack of floating-point support in PDP hardware \emph{and} enable new optimisations.
Moreover, the P4 dataplane can offer runtime control over which flows are monitored.
%Finally, the choice of single-step algorithms (as opposed to $n$-step or Monte Carlo methods) bounds the amount of per-trace state required for online learning to just the last state-action pair, safeguarding the limited memory of the target devices.
We also design our solution to operate as closely as possible to the P4 pipeline to use and learn from per-packet state, but outside of the main packet path to prevent packet stalls.

@@

%The question we investigate is: can online RL be brought to the dataplane by returning to these computationally simpler methods, to act on locally extracted state?
%Can it be made more efficient \emph{by dataplane hardware}?
%Through this work, \approachshort{}, we can comfortably answer ``yes'' on both counts.
%Through \approachshort{}, w
We show that online RL can be brought to the dataplane by such methods, where it can act on locally extracted state and is made \emph{more efficient by dataplane hardware}.
In particular, we exploit how SmartNIC devices often expose general-purpose compute~\parencite{DBLP:conf/sigcomm/LiuCSKPG19} to provide path-adjacent, on-chip RL in the dataplane (\cref{fig:netro-arch}).
As many of these devices have engineering and development histories which predate P4, general compute beyond P4's limits~\parencite{p4-psa} is surprisingly common.
By executing on spare compute units, we prevent packet stalling and offer quick runtime reconfigurability.
%We look backwards to classical RL techniques (tile-coded policies, single-step RL algorithms), alongside the adaptations introduced by existing in-network ML (such as quantisation).
%Such adaptations unlock further improvements in parallel inference and update processing.
This paper contributes:
\begin{itemize}
	\item An analysis of why in-NIC RL is needed and best-placed to interact with the network, made feasible by classical RL methods and quantisation (\cref{sec:motivation}),
	\item \emph{\approachshort{}}: a general-purpose in-NIC RL agent which scales with allocated device resources to meet latency or throughput demands of network traffic control (\cref{sec:design}),
	\item \emph{ParSa}, a wait-free, parallel RL algorithm to accelerate tile coded policy inference and updates (\cref{alg:parsa}),
	\item In-depth evaluation of how \approachshort{} affects carried dataplane traffic, performs under different policy sizes (simple/complex state), and improves on explicit offloading with a \SI{15}{\texttimes} latency reduction compared to commodity hardware (\SI{21}{\texttimes} for 99.99\nthscript{th} tail latencies) and an order of magnitude improvement in online throughput (\cref{sec:evaluation}).
	\item A description of how \approachshort{} would integrate with state-of-the-art PDP applications to perform fully in-NIC, fast, automated DDoS mitigation (\cref{sec:potential-integrations}).
\end{itemize}

@@

\section{Preliminaries: towards in-NIC RL}\label{sec:motivation}
%\section{Preliminaries: RL for in-network computation}\label{sec:motivation}
By combining recent trends in PDP hardware with insights from the ML community (past and present), we discuss why in-NIC RL is needed and best-placed to interact with the network, and how classical RL and quantisation make this feasible.
Crucially, we focus on low port-density devices.
%We envision that this processing is suited to the network edge or key vantage points, rather than core switches.
These devices' designs make it reasonable to move processing \emph{outside of the packet pipeline}; SmartNICs expose additional cores, and NetFPGAs allow for the synthesis of independent functional units.
%The main benefit of doing so is that core control logic can be moved as close to the device as possible \emph{without impacting packet processing rates}.
The main benefit is that RL logic can be moved onto the NIC \emph{without impacting packet processing}.
%?? Say more directly why this mode of operation is preferable to in-pipeline w/ P4
%As Ethernet speed increases, packet deadlines grow tighter; a \SI{64}{\byte} packet must be produced every \SI{12.8}{\nano\second} at \SI{40}{\giga\bit\per\second}, giving for example a \SI{3.99}{\micro\second} deadline on Netronome (which has \num{312} P4 pipelines).
%In-pipeline execution thus increases the risk of packet drops or stalls.

%However, this requires that there be a reasonably consistent interface and behaviour between device classes.
%This need for consistency underpins the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}, which defines a conceptual model of match-action tables divided into ingress and egress pipelines.
%The PSA presents a sensible lower bound on the device capabilities required to implement a P4 dataplane, but the reality is more complex and interesting.

%?? TODO: trim down. Too ``thesis-y'', rather than ``conference-y''.
%?? Save this elsewhere 'til then?

\subsection{Programmable hardware capabilities}
%The introduction of the P4 language~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} has led to explosive growth in the research community surrounding in-network computation and offloading.
%Providing a single language which is compatible with network devices of many form factors has been instrumental in the development of novel fine-grained traffic measurement approaches~\parencite{DBLP:conf/sigcomm/GuptaHCFRW18,DBLP:conf/sigcomm/ChenFKRR18,DBLP:conf/sosr/GhasemiBR17}.
%However, this requires that there be a reasonably consistent interface and behaviour between device classes.
%This need for consistency underpins the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}, which defines a conceptual model of match-action tables divided into ingress and egress pipelines.
%The PSA presents a sensible lower bound on the device capabilities required to implement a P4 dataplane, but the reality is more complex and interesting.

@@

While P4~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} has led to great interest in network programmability, it requires similar behaviour between device classes.
This is encoded by the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}---a lower bound on the device capabilities required to implement a P4 dataplane.
%We would be remiss to believe that P4 and the PSA define all capabilities that programmable hardware supports.
However, many compatible devices long predate these developments. 
Many-core SoC-based Netronome~\parencite{netronome-smartnic}, NetFPGA SUME~\parencite{DBLP:journals/micro/ZilbermanACM14,DBLP:conf/fpga/IbanezBMZ19}, and other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo} allow virtually arbitrary programs to be specified and executed.
%This pattern extends to other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo}.
Currently, low port-density devices like these are most likely to have specific general-purpose compute and high degrees of parallelism~\parencite{DBLP:conf/sigcomm/LiuCSKPG19}, as they are designed for high-performance offloading and middlebox development.
%Even Intel Tofino ASICs~\parencite{barefoot-intel}, which architecturally mirror the PSA, expose additional matching capabilities and \emph{arithmetic logic unit} (ALU) functions via the Tofino Native Architecture.
%Regardless, floating-point operations key to ML workloads are very rarely supported outside of deep learning accelerators (which lack packet switching functionality).
%Moreover, the limited memory and per-packet timing constraints endemic to these devices make this class of in-NIC offloading challenging.

@@

\emph{Reinforcement learning} (RL) trains an \emph{agent} to choose an optimal sequence of actions in pursuit of a given task~\parencite{RL2E}.
Like most ML methods, RL algorithms use gradients to update the parameters used to approximate a function with \emph{reward measurements} from the environment.
RL's Markov Decision Process structure allows online learning of a state-action map in a model-free way, and can step through local minima when needed compared to ML.
In networking, this allows for learning from on-device state or handling rapidly evolving problems.
RL algorithms are agnostic to the policy approximation used so long as they are differentiable, and are computationally simple.
E.g., Sarsa~\parencite[pp. \numrange{217}{221}]{RL2E} requires only additions and multiplication to learn a policy online.

@@

\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.7\linewidth]{figures/coop}
	\caption{\approachshort{}'s compute strategies scale to fit device capacity according to either latency or throughput needs---\Coopfw{} is the optimal online design. A single \emph{controller} delegates RL computation and updates to many \emph{minion} threads, who operate on independent subtasks.\label{fig:single-and-parallel:parallel}}
\end{figure}

\Cref{fig:netro-arch} outlines our design on Netronome SmartNIC hardware in pursuit of this goal: unused device resources beyond the P4-PSA spec \emph{can and should be used} to drive asynchronous environmental control.
We explain relevant NFP architectural details later in \cref{sec:netronome-platform-fundamentals}.
\approachshort{} communicates with the packet pipeline of a P4 dataplane via \texttt{extern} plugins using \inring{} (state, configuration) and \outring{} (action) messages (\cref{fig:netro-arch}).
Internally, \approachshort{} either has all its cores act independently, or cooperate to solve each task (\cref{fig:single-and-parallel:parallel})---with different latency-throughput benefits.
%We open-source our firmware and control programs for the benefit of the community.\footnote{\textanon{\url{https://github.com/FelixMcFelix/pdp-rl-paper}}{Link redacted for double-blind submission.}}
We open-source our firmware and control programs for the benefit of the community.\footnote{Link redacted during submission.}
%Moving beyond the Netronome platform, we describe how our architecture may be adapted and improved on by bespoke hardware or FPGA-based deployment.

\subsection{Challenges, solutions, and insights}\label{sec:challenges-solutions-and-insights}
\fakepara{Network and execution latency}
State measurement, policy computation, and action installation take time to carry out, including serialisation to MAT-friendly rule formats~\parencite{DBLP:conf/infocom/TaylorT05}.
In the midst of these tasks, system state evolves over time---adding noise to the state-action mapping being learned, harming learning and accuracy~\parencite{DBLP:journals/firai/TravnikMSP18}.
To solve this, we \emph{co-locate all RL agent functions on PDP hardware}.

\fakepara{Lack of floating-point}
PDP hardware, being designed solely for efficient packet processing, lacks floating-point arithmetic support even in more general purpose NPU SmartNICs.
Luckily, the embedded ML literature offers many low-precision floating-point formats~\parencite{bfloat16-blog,DBLP:journals/sigops/XieDMKVZT18,DBLP:conf/nips/SunCCWVSCZG19,DBLP:journals/corr/abs-2007-01530} and fixed-point representations without dynamic range~\parencite{DBLP:conf/usenix/ZhangLX00020,DBLP:conf/nsdi/SapioC0NKKKMPR21}.
The latter set requires only integer arithmetic, and allows us to express and update policy parameters online.
Thus, we use \emph{quantised fixed-point ($Qm.n$) representations of action values}.

\fakepara{Costs of \emph{online} function approximation}
Existing ML works on PDP operate by converting a \emph{pre-trained} model into a suitable representation, such as a binary neural network or chain of MATs.
However, NIC-suitable inference schemes must still be trained offline, even though their data formats eschew floating point.
Additionally, DNN training relies on backpropagation, can require memorizing a sizeable replay buffer for RL, and needs many mini-batches of data for stable training.
To solve this, we employ techniques from more classic RL literature: \emph{tile-coded policies}, which have trivial gradients, and \emph{one-step temporal-difference RL algorithms such as Sarsa~\parencite{RL2E}}.

@@

\subsection{Reconfigurability}
\approachshort{} allows policy design and parameters to be changed at runtime using at most two control packets.
Design changes are used at the end of learning (moving from online to offline), or when trying to train a new policy for another task from the same vantage point.
Parameter changes allow an online agent to become more (or less) adaptive to new data (i.e., after detecting a changepoint in traffic).
This extends to policy data, which may be imported from a pre-trained model and exported via PCIe to the host machine.
Some aspects must be chosen at compile time; bit depth, \Coopfw/\Indfw, and maximum policy sizes.
Choosing a bit depth of \qtylist[list-pair-separator = { or }]{16;8}{\bit} halves/quarters policy memory costs, allowing complex problems to be modelled using more dimensions or fine-grained tiles.

>>>>

\begin{abstract}
Data-driven networking is becoming more capable and widely researched, partly driven by the efficacy of \emph{Deep Reinforcement Learning} (DRL) algorithms.
Yet the complexity of both DRL inference and learning force these tasks to be pushed \emph{away from the dataplane} to hosts, harming latency-sensitive applications.
Online learning of such policies cannot occur in the dataplane, despite being useful techniques when problems evolve or are hard to model.

We present \emph{\approachshort{}}---\approach{}---the first work to bring \emph{online reinforcement learning} to the dataplane.
\approachshort{} makes online learning possible in constrained SmartNIC hardware by returning to classical RL techniques---avoiding neural networks.
Our design allows weak yet highly parallel SmartNIC NPUs to be competitive against commodity x86 hosts, despite having fewer features and slower cores.
Compared to hosts, we achieve a \SI{21}{\texttimes} reduction in \num{99.99}\nthscript{th} tail inference times to \SI{34}{\micro\second}, and \SI{9.9}{\texttimes} improvement in online throughput for real-world policy designs.
In-NIC execution eliminates PCIe transfers, and our asynchronous compute model ensures minimal impact on traffic carried by a co-hosted P4 dataplane.
\approachshort{}'s design scales with additional resources at compile-time to improve upon both decision latency and throughput, and is quickly reconfigurable at runtime compared to reinstalling device firmware.
\end{abstract}

@@

%\vspace{-1em}
\section{Introduction}
\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.65\linewidth]{figures/arch-with-p4}
	\caption{\approachshort{} brings low-latency, online reinforcement learning directly to the dataplane. SoC- and NetFPGA-based SmartNIC devices expose spare compute---making in-situ, asynchronous processing and learning possible alongside P4 dataplanes. Classical RL policy methods are the key to making this feasible.\label{fig:netro-arch}}
\end{figure}
Automatic network optimisation, control, and defence are at last becoming commonplace.
Adaptive techniques such as \emph{reinforcement learning} (RL) have led the charge in data-driven networking, enhancing automatic traffic optimisation~\parencite{DBLP:conf/sigcomm/ChenL0L18,DBLP:conf/sigcomm/MaoNA17}, congestion control~\parencite{DBLP:journals/corr/abs-1910-04054}, adaptive routing~\parencite{DBLP:conf/hotnets/ValadarskySST17,DBLP:conf/conext/GiladSGRS20}, resource management~\parencite{DBLP:conf/hotnets/MaoAMK16}, and packet classification~\parencite{DBLP:conf/sigcomm/LiangZJS19}.
In RL methods, every change and its effects improve future decisions.

@@

In parallel, P4~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} and \emph{programmable dataplane} (PDP) hardware~\parencite{DBLP:journals/micro/ZilbermanACM14, netronome-smartnic, xilinx-alveo, barefoot-intel} have inspired explosive growth and interest in the research community surrounding in-network computation and offloading.
The promise of PDP hardware is that we can move the entire monitoring and analysis stack \emph{into the dataplane itself, and have it evolve to incorporate new approaches}.
%Consider a DDoS mitigation system which could react to dataplane-specific, per-packet state in microseconds, and \emph{start learning again just as quickly} in the face of new data or behaviours.
The P4 ecosystem already presents novel, openly-available, and fine-grained traffic measurement techniques~\parencite{DBLP:conf/sigcomm/GuptaHCFRW18,DBLP:conf/sigcomm/ChenFKRR18,DBLP:conf/sosr/GhasemiBR17}, and its control plane makes it easy to select which flows or packets are monitored in a live network.
As a result, there has been keen interest in executing ML in the dataplane~\parencite{DBLP:conf/hotnets/XiongZ19,DBLP:conf/sigcomm/SanvitoSB18,DBLP:journals/corr/abs-1801-05731,DBLP:journals/corr/abs-2009-02353,langlet-ml-netronome,DBLP:journals/corr/abs-2002-08987} to take advantage of flow or per-packet state that cannot be efficiently processed or extracted at any other location.
These works have shown the value of in-network ML: high-throughput, low latency response to network changes.
While they can exploit on-device state to provide reactive insight, the missing piece of the puzzle is learning and updating these ML analyses online without deferring to another machine in the network.
Training these models online and in-network is an exciting (and challenging) lacuna in the field that \emph{has yet to be addressed}.

It is important to make this feasible; offloading to commodity hosts adds PCIe delays~\parencite{DBLP:journals/corr/abs-2009-02353,DBLP:conf/sigcomm/NeugebauerAZAL018}, but is required due to the complexity of modern ML.
For context, deep neural network training relies on backpropagation, can take vast amounts of offline simulation~\parencite{DBLP:journals/corr/abs-1912-06680}, and needs many minibatches for stability.
These induce high costs for compute, storage, or dedicated accelerators to overcome the batching needed to operate at line-rate---while BrainWave~\parencite{DBLP:conf/isca/FowersOPMLLAHAG18} can reduce batching (and thus tail latencies) by \qty{32}{\texttimes}, inference still takes $\mathcal{O}{\left(\text{\si{\milli\second}}\right)}$~\parencite{Duarte2019}.
Moreover, novel DMA techniques such as \emph{GPUDirect}~\parencite{gpudirect} halve but do not eliminate such PCIe transfers.
The high latencies \emph{caused} by steering and inference harm learning~\parencite{DBLP:journals/firai/TravnikMSP18} and runtime application performance~\parencite{DBLP:journals/corr/abs-1910-04054}.
If we can bring online learning \emph{to the dataplane}, then we can take advantage of rich, local state while minimising this impact on the learned policy.
This would also make it easy to train and prototype agent designs which can learn \emph{as the environment evolves}, or when there is too little data to model and simulate a problem.

@@

In this paper, we enable \emph{online in-NIC learning} by returning to \emph{classical} RL methods.
In particular, we focus on tile coding with algorithms such as Sarsa~\parencite{RL2E}.
%These choices have important benefits for in-NIC execution.
While these functions have lower model capacity, they do not require batches of inputs to learn in a stable way, negating the memory needed to store experience replays, and have simple update and inference logic.
In addition, they have shown promise in other network use cases~\parencite{DBLP:journals/eaai/MalialisK15,DBLP:journals/tnsm/SimpsonRP20}.
%Tile coding in particular admits many optimisations, being an embarrassingly parallel problem.
Using fixed-point arithmetic, we solve the lack of floating-point support in PDP hardware \emph{and} enable new optimisations.
Moreover, the P4 dataplane can offer runtime control over which flows are monitored.
%Finally, the choice of single-step algorithms (as opposed to $n$-step or Monte Carlo methods) bounds the amount of per-trace state required for online learning to just the last state-action pair, safeguarding the limited memory of the target devices.
We also design our solution to operate close to the P4 pipeline to access per-packet state, but outside of the main packet path to prevent packet stalls.

@@

%The question we investigate is: can online RL be brought to the dataplane by returning to these computationally simpler methods, to act on locally extracted state?
%Can it be made more efficient \emph{by dataplane hardware}?
%Through this work, \approachshort{}, we can comfortably answer ``yes'' on both counts.
%Through \approachshort{}, w
We show that online RL can be brought to the dataplane by such methods, where it can act on locally extracted state and is made \emph{more efficient by dataplane hardware}.
In particular, we exploit how SmartNIC devices often expose general-purpose compute~\parencite{DBLP:conf/sigcomm/LiuCSKPG19} to provide path-adjacent, on-chip RL in the dataplane (\cref{fig:netro-arch}).
As many of these devices have engineering and development histories which predate P4, general compute beyond P4's limits~\parencite{p4-psa} is surprisingly common.
By executing on spare compute units, we prevent packet stalling and offer quick runtime reconfigurability.
%We look backwards to classical RL techniques (tile-coded policies, single-step RL algorithms), alongside the adaptations introduced by existing in-network ML (such as quantisation).
%Such adaptations unlock further improvements in parallel inference and update processing.
This paper contributes:
\begin{itemize}
%	\item An analysis of why in-NIC RL is needed and best-placed to interact with the network, made feasible by classical RL methods and quantisation (\cref{sec:motivation}),
	\item \emph{\approachshort{}}: a general-purpose in-NIC RL agent which scales with allocated device resources to meet latency or throughput demands of network control (\cref{sec:design}),
	\item \emph{ParSa}, a wait-free, parallel RL algorithm to accelerate tile coded policy inference and updates (\cref{alg:parsa}),
	\item In-depth evaluation of how \approachshort{} affects carried dataplane traffic, performs under different policy sizes, and improves on host deployment with a \SI{15}{\texttimes} latency reduction compared to commodity hardware (\SI{21}{\texttimes} for 99.99\nthscript{th} tail latencies) and an order of magnitude improvement in online throughput (\cref{sec:evaluation}).
	\item A description of how \approachshort{} would integrate with state-of-the-art PDP applications to perform fully in-NIC, automated DDoS mitigation (\cref{sec:potential-integrations}).
\end{itemize}

@@

\section{Preliminaries: towards in-NIC RL}\label{sec:motivation}
%\section{Preliminaries: RL for in-network computation}\label{sec:motivation}
%By combining recent trends in PDP hardware with insights from the ML community, we discuss why in-NIC RL is needed and best-placed to interact with the network, and how classical RL and quantisation make this feasible.
%Crucially, we focus on low port-density devices.
%These devices' designs make it reasonable to move processing \emph{outside of the packet pipeline}; SmartNICs expose additional cores, and NetFPGAs allow for the synthesis of independent functional units.
%The main benefit is that RL logic can be moved onto the NIC \emph{without impacting packet processing}.
%The main benefit of doing so is that core control logic can be moved as close to the device as possible \emph{without impacting packet processing rates}.
%The main benefit is that RL logic can be moved onto the NIC \emph{without impacting packet processing}.
%We envision that this processing is suited to the network edge or key vantage points, rather than core switches.
%?? Say more directly why this mode of operation is preferable to in-pipeline w/ P4
%As Ethernet speed increases, packet deadlines grow tighter; a \SI{64}{\byte} packet must be produced every \SI{12.8}{\nano\second} at \SI{40}{\giga\bit\per\second}, giving for example a \SI{3.99}{\micro\second} deadline on Netronome (which has \num{312} P4 pipelines).
%In-pipeline execution thus increases the risk of packet drops or stalls.
%However, this requires that there be a reasonably consistent interface and behaviour between device classes.
%This need for consistency underpins the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}, which defines a conceptual model of match-action tables divided into ingress and egress pipelines.
%The PSA presents a sensible lower bound on the device capabilities required to implement a P4 dataplane, but the reality is more complex and interesting.
%?? TODO: trim down. Too ``thesis-y'', rather than ``conference-y''.
%?? Save this elsewhere 'til then?
\subsection{Programmable hardware capabilities}
%The introduction of the P4 language~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} has led to explosive growth in the research community surrounding in-network computation and offloading.
%Providing a single language which is compatible with network devices of many form factors has been instrumental in the development of novel fine-grained traffic measurement approaches~\parencite{DBLP:conf/sigcomm/GuptaHCFRW18,DBLP:conf/sigcomm/ChenFKRR18,DBLP:conf/sosr/GhasemiBR17}.
%However, this requires that there be a reasonably consistent interface and behaviour between device classes.
%This need for consistency underpins the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}, which defines a conceptual model of match-action tables divided into ingress and egress pipelines.
%The PSA presents a sensible lower bound on the device capabilities required to implement a P4 dataplane, but the reality is more complex and interesting.

@@

While P4~\parencite{DBLP:journals/ccr/BosshartDGIMRSTVVW14} has led to great interest in network programmability, it requires similar behaviour between device classes---this is encoded by the \emph{Programmable Switch Architecture} (PSA)~\parencite{p4-psa}. %---a lower bound on the device capabilities required to implement a P4 dataplane.
%We would be remiss to believe that P4 and the PSA define all capabilities that programmable hardware supports.
However, many compatible devices long predate these developments. 
Many-core SoC-based Netronome~\parencite{netronome-smartnic}, NetFPGA SUME~\parencite{DBLP:journals/micro/ZilbermanACM14,DBLP:conf/fpga/IbanezBMZ19}, and other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo} allow arbitrary programs to be specified and executed.
%This pattern extends to other SmartNICs~\parencite{nvidia-bluefield,xilinx-alveo}.
Currently, low port-density devices like these are most likely to have general-purpose compute and high degrees of parallelism~\parencite{DBLP:conf/sigcomm/LiuCSKPG19}, as they are designed for high-performance offloading and middlebox development.
%Even Intel Tofino ASICs~\parencite{barefoot-intel}, which architecturally mirror the PSA, expose additional matching capabilities and \emph{arithmetic logic unit} (ALU) functions via the Tofino Native Architecture.
%Regardless, floating-point operations key to ML workloads are very rarely supported outside of deep learning accelerators (which lack packet switching functionality).
%Moreover, the limited memory and per-packet timing constraints endemic to these devices make this class of in-NIC offloading challenging.

@@

\emph{Reinforcement learning} (RL) trains an \emph{agent} to choose an optimal sequence of \emph{actions} from any \emph{state} in pursuit of a given task~\parencite{RL2E}.
Like most ML methods RL uses gradients to update the parameters used to approximate a function, aided by \emph{reward measurements} from the environment.
RL's MDP structure allows online learning of a state-action map in a model-free way, and can step through local minima when needed compared to ML.
In networking, this allows for learning from on-device state or handling rapidly evolving problems.
RL algorithms are agnostic to the policy approximation used so long as they are differentiable, and are computationally simple.
E.g., Sarsa~\parencite[pp. \numrange{217}{221}]{RL2E} requires only additions and multiplication to learn a policy online.

@@

\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.65\linewidth]{figures/coop}
	\caption{\approachshort{} scales to fit device capacity according to either latency or throughput needs---\Coopfw{} is the optimal online design. A single \emph{controller} delegates RL inference and updates to many \emph{minions}, who complete independent subtasks.\label{fig:single-and-parallel:parallel}}
\end{figure}

\Cref{fig:netro-arch} outlines our design on Netronome SmartNIC hardware in pursuit of this goal: unused device resources beyond the P4-PSA spec are used to drive asynchronous environmental control.
We explain relevant NFP architectural details later in \cref{sec:netronome-platform-fundamentals}.
\approachshort{} communicates with the packet pipeline of a P4 dataplane via \texttt{extern} plugins using \inring{} (state, configuration) and \outring{} (action) messages (\cref{fig:netro-arch}).
Internally, \approachshort{} either has all its cores act independently, or cooperate to solve each task (\cref{fig:single-and-parallel:parallel})---with different latency-throughput benefits.
%We open-source our firmware and control programs for the benefit of the community.\footnote{\textanon{\url{https://github.com/FelixMcFelix/pdp-rl-paper}}{Link redacted for double-blind submission.}}
We open-source our firmware and control programs for the benefit of the community~\parencite{pdp-rl-github}.
%Moving beyond the Netronome platform, we describe how our architecture may be adapted and improved on by bespoke hardware or FPGA-based deployment.

\subsection{Challenges, solutions, and insights}\label{sec:challenges-solutions-and-insights}
\fakepara{Network and execution latency}
State measurement, policy inference, and action installation take time, including network latency and serialisation to MAT-friendly formats~\parencite{DBLP:conf/infocom/TaylorT05}.
In the midst of these tasks, system state evolves over time---adding noise to the state-action mapping being learned, harming learning and accuracy~\parencite{DBLP:journals/firai/TravnikMSP18}.
To solve this, we \emph{co-locate all RL agent functions on PDP hardware}.

\fakepara{Lack of floating-point}
PDP hardware, being designed solely for efficient packet processing, lacks floating-point arithmetic support even in more general purpose NPU SmartNICs.
Luckily, the embedded ML literature offers many low-precision floating-point formats~\parencite{bfloat16-blog,DBLP:journals/sigops/XieDMKVZT18,DBLP:conf/nips/SunCCWVSCZG19,DBLP:journals/corr/abs-2007-01530} and fixed-point representations~\parencite{DBLP:conf/usenix/ZhangLX00020,DBLP:conf/nsdi/SapioC0NKKKMPR21}.
The latter set requires only integer arithmetic, which allows us to express and update policy parameters in-NIC.
Thus, we use \emph{quantised fixed-point ($Qm.n$) representations of action values}.

\fakepara{Costs of \emph{online} function approximation}
NIC-suitable inference schemes must still be trained offline, even though their data formats eschew floating-point---existing PDP ML converts a \emph{pre-trained} model into a suitable representation, such as a binary neural network or chain of MATs.
Additionally, DNN training relies on backpropagation, can require memorizing a sizeable replay buffer for RL, and needs many mini-batches of data for stable training.
To solve this, we employ techniques from more classic RL literature: \emph{tile-coded policies}, which have trivial gradients, and \emph{one-step temporal-difference RL algorithms such as Sarsa~\parencite{RL2E}}.

@@

\subsection{Reconfigurability}
\approachshort{} allows policy design and parameters to be changed at runtime using at most two control packets.
Design changes are used at the end of learning (moving from online to offline), or when trying to train a new policy for another task from the same vantage point.
Parameter changes allow an online agent to become more (or less) adaptive to new data (i.e., after detecting a changepoint in traffic).
This extends to policy data, which may be imported from a pre-trained model and exported via PCIe to the host machine.
Some aspects must be chosen at compile time; bit depth, \Coopfw/\Indfw, and maximum policy sizes.
Choosing a bit depth of \qtylist[list-pair-separator = { or }]{16;8}{\bit} halves/quarters policy memory costs, allowing complex problems to be modelled using more dimensions or fine-grained tiles.
In our implementation, configuration packets are carried over UDP and signalled to the dataplane using a reserved DSCP value~\parencite{rfc2474,DBLP:conf/isca/LiLYCSH19}.
While this simplifies parser generation, it also allows for configuration to be received from only trusted hosts (over the dataplane if needed) via P4 rules.
Our control library and evaluation frameworks are written in Rust.
